{memes}: an R interface to the MEME Suite

### 2020-07-23 manuscript.Rmd
change pkg name to {memes}

---
title: '{memer}: an R interface to the MEME Suite'
author:
- ORCID: null
  name: Spencer L. Nystrom
- name: Daniel J. McKay
output: word_document
---

@@

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis including *de-novo* motif discovery, enrichment
testing, comparison, and matching. Here we present `memer`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

`memer` uses standard R and Bioconductor data types for better integration with
common analysis tools like the `tidyverse` and `GRanges`. `memer` outputs also
function as inputs to other `memer` commands, allowing simple construction of motif
analysis pipelines.

In addition to wrapping MEME Suite utilities, `memer` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression. `memer` also provides
simple utilities for data visualization.

`memer` is designed for maximum flexibility and ease of use to allow rapid
iteration during analysis. Here we present several examples of how `memer`
allows novel analyses of ChIP and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R landscape.

@@

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, enrichment
testing,comparison, and matching. Here we present `memer`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

`memer` uses base R and Bioconductor data types for better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memer` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines.

In addition to wrapping MEME Suite utilities, `memer` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression in the tissue
of interest. `memer` also provides simple utilities for data visualization where
appropriate.

`memer` is designed for maximum flexibility and ease of use to allow users to
iterate rapidly during analysis. Here we present several examples of how `memer`
allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R package landscape.

@@

Data from MEME Suite tools can be underutilized because HTML interface?
 - HTML is great human-readable reporting interface
 - `memer` allows customized reporting.

Bioconductor is deprecating PWMEnrich (~AME) and MotIV (~TomTom), `memer` can fill the gap and add additional value.

@@

The key underlying design principle of `memer` was to create tools that easily
integrate into the existing suite of tools available in `Bioconductor` and the
broader `R` package landscape. In this way, `memer` tools can be used to extend
existing analysis pipelines, or to create novel analyses.

@@

At its most basic form, `memer` acts as a wrapper of these functions: users can
point `memer` tools to files on disk, `memer` will construct the commandline
call, output files will be saved on disk and imported into R for further analysis.
However, `memer` is more than just a simple wrapper of the MEME Suite, it is an
analysis framework allowing seamless integration of MEME Suite tools by
providing an abstraction layer which handles file input/output (I/O) to allow
inputs and outputs in the form of in-memory R objects.

When running a MEME Suite tool using `memer`, inputs are allowed in multiple
formats providing the user additional flexibility when performing motif analysis
(see "Core Utilities" for details). Of note, most outputs of `memer` tools can
be used as inputs to other `memer` tools, allowing easy construction of
pipelines (for example, discovering *de-novo* motifs with DREME, then finding
the genomic positions of those motifs in target regions using FIMO).
 - unlike current MEME tools (ex tomtom -> FIMO is hard to do)

@@

Whenever possible, `memer` uses standard R data types for better integration
with common analysis tools, like `dplyr`, `GenomicRanges`, or base R. In this
way, users don't need to learn new ways of working with data, and outputs can
be integrated with other tools with limited effort.

@@

The core functionality of `memer` revolves around the interface to MEME Suite
tools. Each supported MEME Suite tool is assigned a `run` function
(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
accept sequence and motif inputs as needed and any valid commandline flags as
additional function parameters.

@@

Tools requiring sequences as input can use a path to a .fasta format
file, a `Biostrings::XStringSet`, or a `Biostrings::BStringSetList` as input.
These data structures support all biological sequence types: DNA,
RNA, and amino-acid. For the remainder of the text, we will 
**use chip-seq data analysis as a framework to highlight memer features**
\strikethrough{focus our examples on DNA sequences.}

Some `memer` tools require motifs as input. In these instances, `memer` will
always accept a path to a .meme format file, or a list of motifs in
`universalmotif` format \cite{universalmotif}. The `universalmotif` package
provides simple import and motif conversion functions for all common motif data
structures and file types, ensuring cross-compatability with all commonly used
motif data formats. When `memer` functions return motifs, they are returned to
the user in `universalmotif` format. This ensures that all `memer` motif outputs
can be used as inputs to other `memer` functions that accept motif inputs.
Furthermore, `universalmotif` conversion tools allow `memer` detected motifs to
be used in many other popular Bioconductor motif packages, such as `TFBSTools`
and `motifStack`, providing the user with maximum flexibility in their analysis.

@@

A key difference between `memer` and the MEME Suite tools is that `memer`
provides tools allowing the peak-level metadata to frame subsequent analysis.
For DNA ranges (such as in ChIP-seq), the `get_sequence()` utility acts as the
entrypoint to analysis with `memer`. `get_sequence()` returns a `Biostrings`
object containing the DNA sequence of each feature in the input ranges. If
ranges are split into groups by a metadata value, `get_sequence()` will return a
list of sequences for each group. When a sequence list is used as input into
`memer` core functions, the function will run once for each group, enabling a
native "split-apply-combine" workflow without any additional user overhead
\cite{splitapplycombine, wickham}. Together, these features empower users to
use range-based metadata as the centerpoint of their analyses.

### Discriminative analysis
A powerful feature of several MEME Suite Tools is the ability to perform
discriminative analysis, where motifs are identified or tested for enrichment
in one group of sequences relative to another. In many instances, running tools
in discriminative mode on a set of control sequences outperforms simply using
shuffled input sequences as the control (back this up, change wording? maybe:
using biologically relevant control sequences can improve quality of results?).
Therefore, enabling complex discriminative analysis designs was central to the
design of `memer`.

When using list input to `memer` tools, names of list object members can be
passed to the `control` argument. This results in running the tool in
discriminative mode using those sequences as the control sequences for all other
list members. By passing multiple names to `control`, the sequences are combined
together to create a merged background set. In this way, performing data-driven
discriminative analysis is just as simple as one using shuffled sequences as
background.

@@

## Metadata manipulation
`memer` is built on top of the `universalmotif` framework, which provides
methods for importing motifs from various file formats into a unified R data
representation complete with metadata information \cite{universalmotif}.
However, a major limitation of the `universalmotif` framework is that 
manipulating or filtering multiple `universalmotif` entries requires potentially
unfamiliar overhead for less experienced users, acting as a barrier to
performing potentially useful operations.

The `as_universalmotif_dataframe()` function from `memer` converts
`universalmotif` objects into a more familiar `data.frame` format which can be
manipulated using `tidyverse` or base R tools. These data.frames also contain a
special `motif` column which stores the `universalmotif` object of each entry.
Users can update columns in the data.frame which will be reflected in the
corresponding metadata entries of the `motif` by running `update_motifs()` after
manipulating the columns. These data.frames can be converted back into
`universalmotif` format using `as_universalmotif`, which also updates the
metadata using column values. The results from running the core MEME Suite
functions are stored in this format whenever appropriate, creating a consistent
framework for users to explore and manipulate motifs and their associated data values.

@@

# Example of renaming motifs w/ memer syntax
my_motifs <- list(m1,m2)
my_motifs %<>% 
  as_universalmotif_dataframe() %>% 
  dplyr::mutate(name = my_names) %>% 
  as_universalmotif()
```

@@

The MEME Suite provides a limited set of data visualizations which have limited
customizability. `memer` leverages the advantages of the R graphics environment
to provide novel data visualizations which are infinitely customizable by the user.

@@

# Conclusion
 - `memer` is cool
 - use `memer`
 - it's on github (provide link)
 - manual is here:

@@

**Example of commandline vs `memer`?**
For example, to run DREME on the commandline on "input.fasta", while setting the random seed to 100:
```{bash, eval=F}
# Shell command setting random seed to 100
dreme -s 100 input.fasta
```

The `memer` implementation of the above command would be written as follows:
```{r, eval=F}
# memer command setting random seed to 100
# user explicitly states they want to shuffle input sequences as control
runDreme("input.fasta", control = "shuffle", s = 100)
```

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/memer_model.png}
\caption{A diagram of memer tools & workflows?
flowchart of each step, getsequence-> run-> import -> visualize -> goto run

Diagram how meme-suite tools work vs memer?
}
\label{fig:model} 
\end{figure}

>>>>

---
title: '{memes}: an R interface to the MEME Suite'
author:
- ORCID: null
  name: Spencer L. Nystrom
- name: Daniel J. McKay
output: word_document
---

@@

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis including *de-novo* motif discovery, enrichment
testing, comparison, and matching. Here we present `memes`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

`memes` uses standard R and Bioconductor data types for better integration with
common analysis tools like the `tidyverse` and `GRanges`. `memes` outputs also
function as inputs to other `memes` commands, allowing simple construction of motif
analysis pipelines.

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression. `memes` also provides
simple utilities for data visualization.

`memes` is designed for maximum flexibility and ease of use to allow rapid
iteration during analysis. Here we present several examples of how `memes`
allows novel analyses of ChIP and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R landscape.

@@

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, enrichment
testing,comparison, and matching. Here we present `memes`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

`memes` uses base R and Bioconductor data types for better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines.

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression in the tissue
of interest. `memes` also provides simple utilities for data visualization where
appropriate.

`memes` is designed for maximum flexibility and ease of use to allow users to
iterate rapidly during analysis. Here we present several examples of how `memes`
allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R package landscape.

@@

Data from MEME Suite tools can be underutilized because HTML interface?
 - HTML is great human-readable reporting interface
 - `memes` allows customized reporting.

Bioconductor is deprecating PWMEnrich (~AME) and MotIV (~TomTom), `memes` can fill the gap and add additional value.

@@

The key underlying design principle of `memes` was to create tools that easily
integrate into the existing suite of tools available in `Bioconductor` and the
broader `R` package landscape. In this way, `memes` tools can be used to extend
existing analysis pipelines, or to create novel analyses.

@@

At its most basic form, `memes` acts as a wrapper of these functions: users can
point `memes` tools to files on disk, `memes` will construct the commandline
call, output files will be saved on disk and imported into R for further analysis.
However, `memes` is more than just a simple wrapper of the MEME Suite, it is an
analysis framework allowing seamless integration of MEME Suite tools by
providing an abstraction layer which handles file input/output (I/O) to allow
inputs and outputs in the form of in-memory R objects.

When running a MEME Suite tool using `memes`, inputs are allowed in multiple
formats providing the user additional flexibility when performing motif analysis
(see "Core Utilities" for details). Of note, most outputs of `memes` tools can
be used as inputs to other `memes` tools, allowing easy construction of
pipelines (for example, discovering *de-novo* motifs with DREME, then finding
the genomic positions of those motifs in target regions using FIMO).
 - unlike current MEME tools (ex tomtom -> FIMO is hard to do)

@@

Whenever possible, `memes` uses standard R data types for better integration
with common analysis tools, like `dplyr`, `GenomicRanges`, or base R. In this
way, users don't need to learn new ways of working with data, and outputs can
be integrated with other tools with limited effort.

@@

The core functionality of `memes` revolves around the interface to MEME Suite
tools. Each supported MEME Suite tool is assigned a `run` function
(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
accept sequence and motif inputs as needed and any valid commandline flags as
additional function parameters.

@@

Tools requiring sequences as input can use a path to a .fasta format
file, a `Biostrings::XStringSet`, or a `Biostrings::BStringSetList` as input.
These data structures support all biological sequence types: DNA,
RNA, and amino-acid. For the remainder of the text, we will 
**use chip-seq data analysis as a framework to highlight memes features**
\strikethrough{focus our examples on DNA sequences.}

Some `memes` tools require motifs as input. In these instances, `memes` will
always accept a path to a .meme format file, or a list of motifs in
`universalmotif` format \cite{universalmotif}. The `universalmotif` package
provides simple import and motif conversion functions for all common motif data
structures and file types, ensuring cross-compatability with all commonly used
motif data formats. When `memes` functions return motifs, they are returned to
the user in `universalmotif` format. This ensures that all `memes` motif outputs
can be used as inputs to other `memes` functions that accept motif inputs.
Furthermore, `universalmotif` conversion tools allow `memes` detected motifs to
be used in many other popular Bioconductor motif packages, such as `TFBSTools`
and `motifStack`, providing the user with maximum flexibility in their analysis.

@@

A key difference between `memes` and the MEME Suite tools is that `memes`
provides tools allowing the peak-level metadata to frame subsequent analysis.
For DNA ranges (such as in ChIP-seq), the `get_sequence()` utility acts as the
entrypoint to analysis with `memes`. `get_sequence()` returns a `Biostrings`
object containing the DNA sequence of each feature in the input ranges. If
ranges are split into groups by a metadata value, `get_sequence()` will return a
list of sequences for each group. When a sequence list is used as input into
`memes` core functions, the function will run once for each group, enabling a
native "split-apply-combine" workflow without any additional user overhead
\cite{splitapplycombine, wickham}. Together, these features empower users to
use range-based metadata as the centerpoint of their analyses.

### Discriminative analysis
A powerful feature of several MEME Suite Tools is the ability to perform
discriminative analysis, where motifs are identified or tested for enrichment
in one group of sequences relative to another. In many instances, running tools
in discriminative mode on a set of control sequences outperforms simply using
shuffled input sequences as the control (back this up, change wording? maybe:
using biologically relevant control sequences can improve quality of results?).
Therefore, enabling complex discriminative analysis designs was central to the
design of `memes`.

When using list input to `memes` tools, names of list object members can be
passed to the `control` argument. This results in running the tool in
discriminative mode using those sequences as the control sequences for all other
list members. By passing multiple names to `control`, the sequences are combined
together to create a merged background set. In this way, performing data-driven
discriminative analysis is just as simple as one using shuffled sequences as
background.

@@

## Metadata manipulation
`memes` is built on top of the `universalmotif` framework, which provides
methods for importing motifs from various file formats into a unified R data
representation complete with metadata information \cite{universalmotif}.
However, a major limitation of the `universalmotif` framework is that 
manipulating or filtering multiple `universalmotif` entries requires potentially
unfamiliar overhead for less experienced users, acting as a barrier to
performing potentially useful operations.

The `as_universalmotif_dataframe()` function from `memes` converts
`universalmotif` objects into a more familiar `data.frame` format which can be
manipulated using `tidyverse` or base R tools. These data.frames also contain a
special `motif` column which stores the `universalmotif` object of each entry.
Users can update columns in the data.frame which will be reflected in the
corresponding metadata entries of the `motif` by running `update_motifs()` after
manipulating the columns. These data.frames can be converted back into
`universalmotif` format using `as_universalmotif`, which also updates the
metadata using column values. The results from running the core MEME Suite
functions are stored in this format whenever appropriate, creating a consistent
framework for users to explore and manipulate motifs and their associated data values.

@@

# Example of renaming motifs w/ memes syntax
my_motifs <- list(m1,m2)
my_motifs %<>% 
  as_universalmotif_dataframe() %>% 
  dplyr::mutate(name = my_names) %>% 
  as_universalmotif()
```

@@

The MEME Suite provides a limited set of data visualizations which have limited
customizability. `memes` leverages the advantages of the R graphics environment
to provide novel data visualizations which are infinitely customizable by the user.

@@

# Conclusion
 - `memes` is cool
 - use `memes`
 - it's on github (provide link)
 - manual is here:

@@

**Example of commandline vs `memes`?**
For example, to run DREME on the commandline on "input.fasta", while setting the random seed to 100:
```{bash, eval=F}
# Shell command setting random seed to 100
dreme -s 100 input.fasta
```

The `memes` implementation of the above command would be written as follows:
```{r, eval=F}
# memes command setting random seed to 100
# user explicitly states they want to shuffle input sequences as control
runDreme("input.fasta", control = "shuffle", s = 100)
```

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/memes_model.png}
\caption{A diagram of memes tools & workflows?
flowchart of each step, getsequence-> run-> import -> visualize -> goto run

Diagram how meme-suite tools work vs memes?
}
\label{fig:model} 
\end{figure}
